Manuscript title here

### 2024-09-11 manuscript.Rmd
lagt till ofi i pr. typ färdig med pr.

  (4  Kliniskt för att förbättra omhändertagande används OFI
    Förklara vad OFI är för något
      To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI). OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the structure and clinical process of trauma care(Hussein).)

@@

Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used. The classification are divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However this only suits hypovolemic shocks and not the other types. There are two other classifications, namely RETTS and NEWS2. These are used in emergency triaging and can be used on other types of shocks, however they are not shock specific classifications. If we look into all three classifications we will be able to find similarities in that all use HF, BP, RR and mental status. Therefore we decided to use these markers as the independent variables for classification of shock regardless of their nature. (sista två meningarna borde tillhöra metod...)

@@

The outcome is OFI recorded in the trauma care quality database and are binary (as in yes exist or no does not exist) when used to calculate the association with degree of shock. 
The subgroups of OFI will be explored and reported in a list, and therefore non-binary.

To classify the patients the following variables will be used: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.

>>>>

  (4  Kliniskt för att förbättra omhändertagande används OFI
    Förklara vad OFI är för något
      To be able to assess and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI). OFI is multidiciplinary morbidity and mortality reviews, where opportunity for improvements are identified in the structure and clinical process of trauma care[@albaaj_patient_2023]. )

@@

Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used. The classification are divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However this only suits hypovolemic shocks and not the other types. There are two other classifications, namely RETTS and NEWS2. These are used in emergency triaging and can be used on other types of shocks, however they are not shock specific classifications. If we look into all three classifications we will be able to find similarities in that all use HF, BP, RR and mental status. (sista två meningarna borde tillhöra metod...)

@@

To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause. Heart rate, bloodpressure, respiratory rate and mental status. Based on these four measurements, the patients will be calssified into four groups rouphly based on the same value as in the ATLS trauma shock classifications. The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock. The subgroups of OFI will be explored and reported in a list, and therefore non-binary.

### 2024-09-11 project-plan.Rmd
lagt till ofi i pr. typ färdig med pr.

Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year. Among those affected, younger populations are disproportionately impacted, making trauma management a critical issue in healthcare systems worldwide[@noauthor_injuries_nodate]. The initial management of trauma patients in the hospital setting is extremely time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010]. Trauma patients presenting in shock have a notably higher mortality rate compared to those who are not in shock, underscoring the need for improved care strategies for these high-risk patients[@vang_shock_2022].
Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023]. There are four primary causes of shock: hypovolemic, obstructive, cardiogenic, and distributive. These etiologies can occur in isolation or in combination. In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].

To classify shock in trauma 
The Advanced Trauma Life Support (ATLS) system provides a widely used classification for hypovolemic shock based on blood loss, heart rate, blood pressure, respiratory rate, mental status, and urine output. ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe, and helps guide resuscitative efforts. However, this classification is specific to hypovolemic shock and does not account for the other types of shock.

To address the broader spectrum of shock types encountered in trauma, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are commonly used in emergency care settings. While these systems are not shock-specific, they offer useful frameworks for assessing patients in various shock states by prioritizing vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR), and mental status.

RETTS is designed for rapid triage and treatment prioritization in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency. It incorporates physiological parameters similar to those used in ATLS, making it applicable to patients in various forms of shock, including hypovolemic, cardiogenic, and distributive shock.

NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of hospitalized patients. Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature. NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock.

Despite their differences, all three systems—ATLS, RETTS, and NEWS2—share key similarities, particularly in their use of vital sign markers such as heart rate, blood pressure, respiratory rate, and mental status. This overlap suggests a potential framework for integrating these systems to better classify trauma patients presenting in any form of shock.

Infoga OFI

>>>>

Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year. Among those affected, younger populations are more impacted, making trauma management a worldwide health care concern[@noauthor_injuries_nodate]. The initial management of trauma patients in the hospital setting is time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010]. Trauma patients presenting with shock have a notably higher mortality compared to those who are not in shock, underscoring the need for improved care strategies for these patients[@vang_shock_2022].
Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023]. There are four primary causes of shock: hypovolemic, obstructive, cardiogenic and distributive. These etiologies can occur alone or in combination. In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].

To classify shock in trauma 
The Advanced Trauma Life Support (ATLS) classification for hypovolemic shock based on the parameters of "blood loss", "heart rate", "blood pressure", "respiratory rate", "mental status" and "urine output". ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe. However, this classification is specific to hypovolemic shock and does not account for the other types of shock like obstructive.

To address other types of diseases, such as other types of trauma shock, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are used in emergency care. While these systems are not shock-specific, they are useful for assessing patients in various shock states by using vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR) and mental status.

RETTS is designed for triage in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency. It incorporates parameters similar to those used in ATLS, but also include airway, saturation and temperature. This makes it applicable to patients in various forms of shock, including hypovolemic, cardiogenic and distributive shock(källa?).

NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of patients. Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature. NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock(källa?).

Despite their differences, all three systems ATLS, RETTS and NEWS2 share similarities in their use of vital sign markers such as heart rate, blood pressure, respiratory rate and mental status. This overlap suggests using these four parameters as a base for general classification for all types of shock is justifiable(är det verkligen så, får man anta det?).

To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases. Multidisciplinary morbidity and mortality reviews play a central role in this process, where clinicians from different specialities together analyse trauma cases to evaluate both the structure and clinical processes involved in trauma care following the Donabedian’s quality of care framework[@albaaj_patient_2023]. For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa). By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity(källa).
